---
title: "STAT-S675 HDA_Assignment 7"
author: "Rahul Raghatate"
date: "October 11, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,fig.width=10, fig.height=8)
```

```{r include=FALSE, cache=FALSE}
source("stress.R")
source("manifold.R")
```
# Q1

GIVEN:\
$y_i = (\cos \theta_i,\sin \theta_i) \in R^2$ where,\
$\theta = (i-1) \pi/3$ for i = 1, . . . , 6.
```{r fig_width=6,fig_height=3}
y = matrix(0, 6, 2)
for(i in 1:6) {
  theta = (i-1) * (pi/3)
  y[i, 1] = cos(theta)
  y[i, 2] = sin(theta)
}
print(y)
plot(y,main="Plot for yi in R^2:")
```


Also given,\
$a = \sqrt(2 - 2*\cos(\pi/3))$ \
$b = \sqrt(2 + 2*\cos(\pi/3))$
```{r fig_width=6,fig_height=4}
a = sqrt(2 - 2*cos(pi/3))
b = sqrt(2 + 2*cos(pi/3))

#Given dissimilarity matrix after computing shortest path dist 
#between vertices of undirected graph

Delta = matrix(c(0, a, b, 2, a+2, b+2,
             a, 0, a, b, 2, a+2,
             b, a, 0, a, b, 2,
             2, b, a, 0, a, b,
             a+2, 2, b, a, 0, a,
             b+2, a+2, 2, b, a, 0), nrow = 6)
```
Constructing an initial configuration by CMDS,
```{r}
X.cmds = cmdscale(Delta)

cat("the raw stress of X.cmds:",
    mds.stress.raw.eq(X.cmds,Delta))
```
Performing 20 Successive guttman iterations,
```{r}
X.cmds.transf<-X.cmds
for (i in 1:20){
 X.cmds.transf<-mds.guttman.eq(X.cmds.transf,Delta)
 if (i%%5==0){
 cat("raw stress of the new configuration after k=",i,"=",
     mds.stress.raw.eq(X.cmds.transf,Delta),"\n")
 }
}
cat("Configuration after obtaining initial CMDS configuration and 20 successive Guttman iterations")
plot(X.cmds.transf,main = "Final Configuration")                        
```

# Q2

Loading the data,
```{r}
X.spiral = read.csv("http://mypage.iu.edu/~mtrosset/Courses/675/X.spiral", 
                    sep = "", header=  FALSE)
plot(X.spiral,main="Orignal Spiral plot")

cat('EDM-1 for X.spiral')
X.spiral.edm1 = mds.edm1(X.spiral)
h_values=c(1e-3,1e-2,0.05,1e-1,0.5,1,1.5,3.5,5,10,20,25,50)
par(mfrow=c(2,2))
for (h in h_values){
    X.spiral.w = graph.heat(X.spiral.edm1,h)
    G = graph.laplacian(X.spiral.w)
    laplacian.eig = eigen(G)
    val = laplacian.eig$values
    val.vec = as.vector(sqrt(val))
    laplacian.eig.vec = laplacian.eig$vectors[, 199:200]
    X.spiral.lp = sweep(laplacian.eig.vec, MARGIN = 2, val.vec[199:200], "/")
    plot(X.spiral.lp,main = paste("Laplacian Eigenmap for h=",h))
}
```
Observations:
* For small values of h in range of (0.001 to 0.1),the configurations looks clustered and the data is not similar to original configuration. The similarities for h<0.1 still seems to be different hence no overlapping.
* As the value of h increases for (h=0.1) the data configurations seems to be a good approximation.
* But,as h increases further till 50,the points seems to overlap, and configuration shows distortion.The points with similarities farther have come closer and hence overlaping can been seen. 

# Q3
```{r}

mds.tau <- function(H)
{
#
#  This function returns the double centering of the inputted matrix.
#  See Critchley for details.
#
        n <- nrow(H)
        P <- diag(n) - 1/n
        return(-0.5 * P %*% H %*% P)
}
mds.kappa <- function(C) {
#
#  Critchley's kappa operator on centered nxn matrices.
#  This is the inverse of tau.
#
        n <- nrow(C)
        H <- matrix(1,nrow=n,ncol=n)
        d <- diag(C)
        H <- diag(d) %*% H
        H <- H + t(H) - 2 * C
        d <- seq(1, n^2, n + 1)
        H[d] <- 0
        return(H)
}

X.spiral = read.csv("http://mypage.iu.edu/~mtrosset/Courses/675/X.spiral", sep = "", header=  FALSE)
#Delta.sp<-matrix(c(0,1.6,1.6,1.6,3,0,3,3,3,3,0,3,1.6,1.6,1.6,0),nrow=4)
#X.spiral<-X.spiral[1:4,]
n = nrow(X.spiral)
Delta.sp = mds.edm1(X.spiral)
h = 0.1
W = exp(-h*(Delta.sp^2))
e = matrix(1, nrow = n)
t = W %*% e
T = diag(as.vector(t))
T_inv = solve(T)
L = T - W
P = T_inv %*% W
w_bar = as.numeric(t(e) %*% W %*% e)
PI = t/w_bar
D = w_bar*T_inv
Z = solve(T_inv%*%L + e%*%t(PI))
Z_D = Z %*% D
# Z_D = solve((L/w_bar) + PI %*% t(PI))
C = mds.kappa(Z_D)
B = mds.tau(C)
B.eig = eigen(B)
B.eig$values
```
From Section  6.4.1-6.4.2 we know that  calculated C in above code is EDM-2. 

To prove $\Delta = [\delta_{ij}]$ as EDM-1:  

* compute $A = \Delta_2 = [\delta^2_{ij}]$  
* compute $B = \tau(A) = -(1/2) PAP$  
* compute the spectral decomposition of $B$  
* if there are any negative eigen values, then $\Delta$ is not EDM-1, otherwise it is EDM-1.

I couldn't find any specific configuration satisfying above conditions. I tested for Q2 data.
