---
title: "Multinomial Regression for Student Program Choice Prediction"
author: "Rahul Raghatate [rraghata@iu.edu]"
date: "October 26, 2017"
output: pdf_document
---


Students entering high school can make program choices among general program, vocational program and academic program.

So, to predict the program type , 

Lets try to fit multinomial logistic regression for their choice of program with respective to other categorical and ordinal variables 

Data: https://stats.idre.ucla.edu/stat/data/hsbdemo.dta

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE,error=FALSE,tidy=TRUE,fig.align='center',fig.width=10,fig.height=6)
#Libraries required
library(nnet)
library(reshape2)
library(ggplot2)
library(broom)
library(tidyr)
library(GGally)
library(gridExtra)
library(foreign)
library(VGAM)
library(MASS)
library(arm)
library(devtools)
library(ggbiplot)
```
## Data Import
Lets import the data and have a look at summary of it.
```{r}
#library(rio)
hsb_data = read.dta("hsbdemo.dta")
#Required predictors and dependent variable
head(hsb_data[,c(3,7,5)])
summary(hsb_data[,c(3,7,5)])
# baseline level of 'prog' dependent variable
hsb_data$prog_l <- relevel(hsb_data$prog, ref = "academic")
```
## Exploring bivariate relation counts
```{r}
# ses ~ prog_l
with(hsb_data, table(ses, prog_l))
# write ~ prog_l
with(hsb_data, do.call(rbind, tapply(write, prog_l, function(x) c(M = mean(x), SD = sd(x)))))
```
## Applying Multinomial Logistic Regression Model
```{r}
#Multinomial Regression
model <- multinom(prog_l ~ ses + write, data = hsb_data)
```
```{r echo=FALSE}
summary(model)
# Store coefficients
coefs<-coef(model)
#transforming for direct interpretation
coefs<-(exp(coefs)-1)*100
#Displaying Coefficients
coefs
```
First row compares prog = "general" to baseline prog = "academic" . Second row compares prog = "vocation" to baseline prog = "academic".

Interpretations for the General vs. Academic model:\

$\ln(\frac{P(prog="general")}{P(prog="academic")})=b_{10}+b_11*(ses=2)+b_{12}*(ses=3)+b_{13}*write$ \
Interpretations for the Vocational vs. Academic model:\

$\ln(\frac{P(prog="vocation")}{P(prog="academic")})=b_{20}+b211*(ses=2)+b_{12}*(ses=3)+b_{23}*write$

## Plot the fitted values against writing score and social economic status
```{r}
model.df <- data.frame(ses=hsb_data$ses, write=hsb_data$write)
model.df.prob <- cbind(model.df, fitted.values(model))
#dataframe for ggplot
pred_prob.ggplot <- melt(model.df.prob,id.vars =
                          c("ses","write"),value.name = "probability")
## plot of fitted probabilities facet over program type
ggplot(pred_prob.ggplot, aes(x = write, y = probability, colour = ses))+
  geom_line() +
  facet_grid(variable ~ ., scales = "free")+
  ggtitle(" variation in probability for each outcome categories")+
  theme(plot.title = element_text(size = 10, face = "bold"))
```
## Based on the models defined, for "ses=middle and write=54", making a prediction for prog
```{r}
data <- data.frame(ses = "middle", write = 54)
cat("\nFor ses=middle and write=54, the predicted probabilities for prog_type are:\n",predict(model,newdata=data,"probs"))
probs<-predict(model,newdata=data,"probs")
#Calculating required ratios

vocation_over_academic=probs[3]/probs[1]

general_over_academic=probs[2]/probs[1]

cat("\n\nProbability of choosing each outcome category over the baseline category are \n")
vocation_over_academic
general_over_academic
```
## Proportional odds logistic regression
Lets treat prog as an ordered categorical variable and fit a proportional odds logistic regression with the same predictors ses and write.
```{r}
#Model
hsb_data.polr =polr(factor(prog) ~ ses+write,data =hsb_data)
display(hsb_data.polr)
probs<-predict(hsb_data.polr,data.frame(ses='middle',write=54),type ="probs")
cat("\nProbability for each outcome category.\n")
probs
```

### Therefore, prediction for prog given ses=middle and write=54 is "academic".

We also have their reading, math, science and standardized test score for social studies in the data set ("read", "math", "science" and "socst").

##  PCA for above variables and "write" to explore first two principal components
```{r}
pca_cols<-c('read','math','science','socst','write')
pca_data<-hsb_data[pca_cols]
data.pca =prcomp(pca_data,scale.=TRUE)
data.pca
ggbiplot(data.pca, obs.scale = 1, groups = hsb_data$prog)+
  ggtitle("Biplot for first two principal components")+
  theme(plot.title = element_text(size = 10, face = "bold"))
```